Topology and experimental distinguishability

### 2017-07-13 Articles/TopologyAndExperimentalDistinguishability/article.tex

\title{Topology and physical distinguishability}

@@

\author[umich]{Christine A. Aidala}
\ead{caidala@umich.edu}

\author[umich]{Gabriele Carcassi\corref{cor1}}
\ead{carcassi@umich.edu}

\author[umich]{Mark J. Greenfield}
\ead{markjg@umich.edu}

\address[umich]{University of Michigan, Ann Arbor, MI 48109, USA}

@@

\begin{abstract}
We formalize the relationship between topological spaces and the ability to distinguish objects experimentally, providing understanding and justification as to why topological spaces and continuous functions are pervasive tools in the physical sciences. The aim is to use these ideas as a stepping stone to give a more rigorous physical foundation to dynamical systems and, in particular, Hamiltonian dynamics.  We first define an experimental observation as a statement that can be verified using an experimental procedure. We show that observations are not closed under negation and countable conjunction, but are closed under finite conjunction and countable disjunction. We then consider observations that identify elements in a set and show how they induce a Hausdorff and second countable topology on that set, thus identifying an open set as one that can be associated with an experimental observation. We then show that only continuous functions preserve experimental distinguishability and that the collection of these functions can be given a Hausdorff and second countable topology. This shows that the universe of discourse of experimental distinguishability so defined is closed and consistent.

@@

The successful use of mathematical ideas in experimental sciences is something long established and celebrated.\footnote{Unreasonable effectiveness and other} In particular topology is arguably the most widespread, as it provides the foundation for other mathematical tools such as differential geometry and ???. Why is it so successful? What physical property is captured by topological spaces that is so fundamental for scientific investigation?

We believe we have identified the answer: topologies capture the way objects are experimentally distinguishable. It is open sets that are associated with experimental outcomes, not the values themselves. For example, as real values are only measurable with finite precisions, the standard topology on euclidean spaces is the one that is physically meaningful. Along the same line, experimental relationships will need to preserve what can be experimentally distinguished and they will therefore be continuous functions.

In this sense, experimentally distinguishability is \emph{the} application of topology in experimental science. It is fitting, then, that topology is so wide-spread as defining what can be experimentally identified is truly a fundamental aspect in science. The aim of this paper is to lay down a formal framework that formalizes this insight.

The work is organized as follows. We will first define experimental observations as statements paired with an experimental test that is able to verify them. We will study their properties under logical operations and show they are only closed under finite conjunction and countable disjunction.

We will then show that any set of objects that are experimentally distinguishable is a Hausdorff second-countable topological space. Both the points and the topology of the space will be formally defined in terms of the observations themselves.

Finally, we will study relationships between experimentally distinguishable objects. This will be represented by continuous functions and can be shown to be themselves physically distinguishable

We believe this work not only provides interest insights, but a firm justification of the use of topological ideas in the physical science.

>>>>

\title{Topology and experimental distinguishability}

@@

\author[]{Christine A. Aidala\corref{cor1}}
\ead{caidala@umich.edu}

\author[]{Gabriele Carcassi}
\ead{carcassi@umich.edu}

\author[]{Mark J. Greenfield}
\ead{markjg@umich.edu}

\address{University of Michigan, Ann Arbor, MI 48109, USA}

@@

\begin{abstract}
We formalize the relationship between topological spaces and the ability to distinguish objects experimentally, providing understanding and justification as to why topological spaces and continuous functions are pervasive tools in science. We first define an experimental observation as a statement that can be verified using an experimental procedure and show that observations are closed under finite conjunction and countable disjunction. We then consider observations that identify elements in a set and show how they induce a Hausdorff and second-countable topology on that set, thus identifying an open set as one that can be associated with an experimental observation. We then show that only continuous functions preserve experimental distinguishability and that the collection of these functions can be given a Hausdorff and second-countable topology.

@@

The successful use of mathematical ideas in experimental sciences is something long established and celebrated.\footnote{Unreasonable effectiveness and other} In particular topology is arguably the most widespread, as it provides the foundation for other mathematical tools such as differential geometry and ???. Why is it so successful? What property is captured by topological spaces that is so fundamental for scientific investigation?

We believe we have identified the answer: topologies capture the way objects are experimentally distinguishable. It is open sets that are associated with experimental outcomes, not the points themselves. For example, as real values are only measurable with finite precision, the standard topology on Euclidean spaces is the one that is physically meaningful. Along the same line, experimental relationships need to preserve what can be experimentally distinguished, and they are therefore represented as continuous functions.

In this sense, experimental distinguishability is \emph{the} application of topology in experimental science. It is fitting, then, that topology is so widespread given that defining what can be experimentally identified is truly a fundamental aspect of science. The aim of this paper is to lay down a framework that formalizes this insight.

The work is organized as follows. We first define experimental observations as statements paired with an experimental test that is able to verify them. We study their properties under logical operations and show that they are only closed under finite conjunction and countable disjunction.  We then show that any set of objects that are experimentally distinguishable is a Hausdorff second-countable topological space. Both the points and the topology of the space are formally defined in terms of the observations themselves.  Finally, we study relationships between experimentally distinguishable objects. These are represented by continuous functions, and they can be shown to be themselves physically distinguishable.

We believe this work not only provides interesting insights, but a firm justification of the use of topological ideas in the sciences.
